###Exploratory Data Analysis by Allan Visochek
####Using the Prosper Loan Company Dataset
#####06-01-2015

The following exploratory data analysis is an 

investigation of the factors influencing the 

borrower rate of loans from the Prosper Loan Company 

from the second quarter of 2006 through the first 

quarter of 2014.


```{r echo=FALSE,packages ,message=FALSE, warning=FALSE}
# Load all of the packages that you end up using
# in your analysis in this code chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk.
# This prevents the code from displaying in the knitted HTML output.
# You should set echo=FALSE for all code chunks in your file.

library(ggplot2)
library(scales)
library(gridExtra)
```

```{r echo=FALSE,read_data,message=FALSE,warning=FALSE}
loanData<-read.csv("/home/allan/Desktop/loan__exploratory_data_analysis_with_R/prosperLoanData.csv")
```

# Univariate Plots Section

### -------------------------------

### BorrowerRate Histogram & Summary

```{r echo=FALSE, Univariate_Plots_1.1,warning=FALSE}
ggplot(aes(x=BorrowerRate),
       data=loanData)+
  geom_histogram(binwidth=.005)+
  scale_x_continuous(breaks=seq(0,.5,.02))+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))

summary(loanData$BorrowerRate)
```

BorrowerRate Ranges from 0 to .05

#### BorrowerRate Quantiles:

```{r echo=FALSE, Univariate_Plots_1.2,warning=FALSE}
quantile(loanData$BorrowerRate,.1)
quantile(loanData$BorrowerRate,.9)
```

Most BorrowerRates are between .099 and .310

### LoanOriginalAmount Histogram and Summary

```{r echo=FALSE, Univariate_Plots_1.3,warning=FALSE}
# loanData$BorrowerRateCategory<-
#   cut(loanData$BorrowerRate,
#       c(0,0.1,0.31,0.5),
#       labels=c('low','normal','high'))


ggplot(aes(x=LoanOriginalAmount),
       data=loanData)+
  geom_histogram(binwidth = 5000)
summary(loanData$LoanOriginalAmount)
```

all loans are between 0 and 35,000...

75% of loans are under 12,000...

### Loan Term, Bar Chart 
```{r echo=FALSE, Univariate_Plots_1.4,warning=FALSE}
loanData$Term <-factor(loanData$Term)
ggplot(aes(x=Term),
       data=loanData)+
  geom_bar()
```

All loans have a term of either 1,3 or 5 years.

### Employment Status Summary

```{r echo=FALSE, Univariate_Plots_1.5,warning=FALSE}
summary(loanData$EmploymentStatus)
```

The large majority of loans go to individuals who have a status of either "Employed" or "Full Time". No surprises here.

### Homeowner Summary

```{r echo=FALSE, Univariate_Plots_1.6,warning=FALSE}
summary(loanData$IsBorrowerHomeowner)
```

About half of the Borrowers are homeowners

### Borrower State Summary

```{r echo=FALSE, Univariate_Plots_1.7,warning=FALSE}
summary(loanData$BorrowerState)
```

There is a pretty even distribution of borrowers among the States, when taking state population into account.

### Occupation Summary

```{r echo=FALSE, Univariate_Plots_1.8,warning=FALSE}
summary(loanData$Occupation)
```

### Prosper Rating Summary

```{r echo=FALSE, Univariate_Plots_1.9,warning=FALSE}
summary(loanData$ProsperRating..Alpha.)
```

### Credit Grade Summary

```{r echo=FALSE,warning=FALSE}
summary(loanData$CreditGrade)
```

There are a large number of Borrowers that don't have a prosper rating or don't have a credit score...

```{r echo=FALSE, Univariate_Plots_1.10,warning=FALSE}
loanData$HasCreditGrade <- 
   !(loanData$CreditGrade==''|is.na(loanData$CreditGrade))
loanData$HasProsperRating <-
   !(loanData$ProsperRating..Alpha.==''|is.na(loanData$ProsperRating..Alpha.))

# I'm also going to switch the order so that AA comes before A
 loanData$CreditGrade <-
   factor(loanData$CreditGrade,
          levels=c("AA","A","B","C","D","E","HR","NC",""))
 loanData$ProsperRating..Alpha. <- 
   factor(loanData$ProsperRating..Alpha.,
         levels =c("AA","A","B","C","D","E","HR",""))
```

### Credit Score Histogram and Summary

```{r echo=FALSE,warning=FALSE}
ggplot(aes(x=CreditScoreRangeLower),data=loanData)+
  geom_bar(binwidth=10)+
  scale_x_continuous(limits=c(440,900),breaks=seq(440,880,20))+
  theme(axis.text.x=element_text(angle = 60, hjust = 1))
summary(loanData$CreditScoreRangeLower)
```

Most borrowers have a credit score between 600 and 800

### IncomeRange Summary

```{r echo=FALSE, Univariate_Plots1.12,warning=FALSE}
summary(loanData$IncomeRange)
```

Few loans are given out to individuals with low income, or who are unemployed, no surprises here.

```{r echo=FALSE, Univariate_Plots1.13,warning=FALSE}
#Let's reorder IncomeRange and create a new variable to indicate weather a borrower is reported as having an income.

loanData$IncomeRange <- 
  factor(loanData$IncomeRange,
         levels=c("$100,000+",
                  "$75,000-99,999",
                  "$50,000-74,999",
                  "$25,000-49,999",
                  "$1-24,999","$0",
                  "Not employed",
                  "Not displayed"))

loanData$HasIncome <-
   loanData$IncomeRange == "$75,000-99,999"|
   loanData$IncomeRange == "$50,000-74,999"|
   loanData$IncomeRange == "$25,000-49,999"|
   loanData$IncomeRange == "$1-24,999"
```

### StatedMonthlyIncome Histogram and Summary

```{r echo=FALSE, warning=FALSE}
# ggplot(aes(StatedMonthlyIncome),
#        data=loanData)+
#   geom_histogram(binwidth=1000)
# 
# ggplot(aes(StatedMonthlyIncome),
#        data=subset(loanData,
#                    StatedMonthlyIncome>0))+
#   geom_histogram(binwidth=.05)+
#   scale_x_log10()

ggplot(aes(StatedMonthlyIncome),
       data=subset(loanData,
                   StatedMonthlyIncome>0))+
  geom_histogram(binwidth=.05)+
  scale_x_log10(limits=c(500,50000),
                breaks=c(1000,3000,10000,30000))
summary(loanData$StatedMonthlyIncome)
```

### DebtToIncomeRatio

```{r echo=FALSE, Univariate_Plots1.14,warning=FALSE}
# ggplot(aes(x=DebtToIncomeRatio),
#        data=loanData)+
#   geom_histogram(binwidth=.05)

ggplot(aes(x=DebtToIncomeRatio),
       data=subset(loanData,
                   !is.na(DebtToIncomeRatio)))+
  #scale_x_log10(breaks=c(.01,.03,.1,.3,.49,1,10))+
  scale_x_log10()+
  coord_cartesian(xlim=c(0.001,1))+
  geom_histogram(binwidth=.05)

# ggplot(aes(x=DebtToIncomeRatio),
#        data=subset(loanData,
#                    !is.na(DebtToIncomeRatio)))+
#   #scale_x_log10(breaks=c(.01,.03,.1,.3,.49,1,10))+
#   coord_cartesian(xlim=c(0,1))+
#   geom_histogram(binwidth=.05)

summary(loanData$DebtToIncomeRatio)
```

Most borrowers have a DebtToIncomeRatio in the range of 0 to 0.5. Those with a ratio of 10 are probably errors.

### Loan Period

```{r echo=FALSE, Univariate_Plots,warning=FALSE}
# ggplot(aes(x=LoanOriginationQuarter),
#        data=loanData)+
#   geom_bar()+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))


#Lets reorder these by time and try again


loanData$LoanOriginationQuarter <- 
  factor(loanData$LoanOriginationQuarter, 
         levels = c("Q1 2006","Q2 2006","Q3 2006","Q4 2006",
                    "Q1 2007","Q2 2007","Q3 2007","Q4 2007",
                    "Q1 2008","Q2 2008","Q3 2008","Q4 2008",
                    "Q1 2009","Q2 2009","Q3 2009","Q4 2009",
                    "Q1 2010","Q2 2010","Q3 2010","Q4 2010",
                    "Q1 2011","Q2 2011","Q3 2011","Q4 2011",
                    "Q1 2012","Q2 2012","Q3 2012",
                    "Q1 2013","Q2 2013","Q3 2013","Q4 2013",
                    "Q1 2014"))

ggplot(aes(x=LoanOriginationQuarter),
       data=subset(loanData,
                   !is.na(LoanOriginationQuarter)))+
  geom_bar()+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

We can see here that the number of loans given out dropped dramatically in the last quarter of 2008, likely due to the recession.

# Univariate Analysis

### ---------------------

### What is the structure of your dataset?
There are 113,937 loans in the dataset with 81 features, 13 of which were used in the analysis: 

#### Numerical Variables:

BorrowerRate

CreditScoreRangeLower

DebtToIncomeRatio

LoanOriginalAmount

StatedMonthlyIncome 

#### Ordered factor variables:
##### (from best to worst / greatest to least...)

##### Term: 
60,36,12

>(note that term was a numerical variable but was transformed to a factor variable because it has very few values)

##### CreditGrade: 
AA,A,B,C,D,E,HR,NC,none

##### ProsperRating..Alpha.: 
AA,A,B,C,D,E,HR,none

##### IncomeRange: 
$75,000-99,999 ; $50,000-74,999 ; $1-49,999 1-24,999 ; $0 ; Not employed; Not displayed

#### Unordered factor variables:

##### EmploymentStatus:  
Employed, Full-time, Not employed, Part-time, Retired, Self-employed none,Not available, Other,

##### Occupation: 
Accountant/CPA, Administrative Assistant, Analyst, Architect, Attorney, Biologist, Bus Driver, Car Dealer, Chemist, Civil Service, Clergy, Computer Programmer, Construction, Dentist, Doctor, Engineer - Chemical, Engineer - Electrical, Engineer - Mechanical, Executive, Fireman, Flight Attendant, Food Service, Food Service Management, Homemaker, Investor, Judge, Laborer, Landscaping, Medical Technician, Military Enlisted, Military Officer, Nurse (LPN), Nurse (RN), Nurse's Aide, Other, Pharmacist, Pilot - Private/Commercial, Police Officer/Correction Officer, Postal Service, Principal, Professional, Professor, Psychologist, Realtor, Religious, Retail Management, Sales - Commission, Sales - Retail, Scientist, Skilled Labor, Social Worker, Student - College Freshman, Student - College Graduate Student, Student - College Junior, Student - College Senior, Student - College Sophomore, Student - Community College, Student - Technical School, Teacher, Teacher's Aide, Tradesman - Carpenter, Tradesman - Electrician, Tradesman - Mechanic, Tradesman - Plumber, Truck Driver, Waiter/Waitress

##### BorrowerState:
AK, AL, AR, AZ, CA, CO, CT, DC, DE, FL, GA, HI, IA, ID, IL, IN, KS, KY, LA, MA, MD, ME, MI, MN, MO, MS, MT, NC, ND, NE, NH, NJ, NM, NV, NY, OH, OK, OR, PA, RI, SC, SD, TN, TX, UT, VA, VT, WA, WI, WV, WY 

#### Other observations.

BorrowerRate Ranges from 0 to .05

Most BorrowRates are between .099 and .310

Loans amounts range from 0 to $35,000. 

75% of loans are for under $12,000.

All loans have a term of either 1,3 or 5 years.

The large majority of loans go to individuals who are employed full time.

A large number of Borrowers either don’t have a prosper rating or don't have a credit score.

Few loans are given out to individuals with low income, or who are unemployed.

The number of loans given out dropped dramatically in the last quarter of 2008

### What is/are the main feature(s) of interest in your dataset?

The main feature of interest for this investigation is the BorrowerRate. Again, this investigation is primarily concerned with the factors influencing the borrower rate.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?

It is hard to say at this point which features will be most useful based solely on the univariate plots above. All of the variables mentioned above were selected for the investigation because they are likely to have an impact on the borrower rate.

### Did you create any new variables from existing variables in the dataset?

HasCreditGrade -> CreditGrade is available

HasProsperRating -> ProsperRating is available

HasIncome -> IncomeRange is available and greater than 0

##### Additional variables are created in later sections

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?

I log transformed StatedMonthlyIncome and DebtToIncomeRatio, which had left leaning distributions:

##### I reordered the following factor variables:

ProsperScore..Alpha,

CreditGrade,

Incomerange,

LoanOrginationQuarter

#### I also subsetted the data for StatedMonthlyIncome by IncomeVerifiable.

# Bivariate Plots Section
### -------------------------------
### Borrower Rate Over Time
```{r echo=FALSE, Bivariate_Plots1.1,warning=FALSE}
ggplot(aes(x=LoanOriginationQuarter,y=BorrowerRate,group=1),
       data=subset(loanData,
                   !is.na(LoanOriginationQuarter)))+
  geom_line(stat='summary',fun.y=median)+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))+
  ylab("Median BorrowerRate")
```

The median Borrower Rate was decreasing significantly from 2012 to 2014. The wide variation in mean BorrowerRate over time means it will be important to facet by quarter in the multivariate plots section.

### BorrowerRate v. Loan Term

```{r,echo=FALSE,warning=FALSE}
ggplot(aes(y=BorrowerRate,x=Term),data=loanData)+
  geom_boxplot()
with(loanData,by(BorrowerRate,Term,summary))
```

1 year loans have a slightly lower rate than the others.  I will take out the 1 and 5 year loans, since most loans have a 3 year term, (as observed in the univariate plots section).

```{r,echo=FALSE,warning=FALSE}
loanData<-subset(loanData,Term=="36")
```

### BorrowerRate v. LoanOriginalAmount

```{r echo=FALSE, warning=FALSE}

ggplot(aes(y=BorrowerRate,x=LoanOriginalAmount),data=loanData)+
  geom_point()
```

In order to get a better picture, I'm going to

cut the LoanOriginalAmount variable into Quantiles.

### Median BorrowerRate v. LoanOriginalAmount Quantile
```{r,echo=FALSE,warning=FALSE}
## cut loanOriginalAmount by quantiles...
loanData$LoanOriginalAmountQuantile <- cut(loanData$LoanOriginalAmount,
                                           with(subset(loanData,!is.na(LoanOriginalAmount)),
                                               quantile(LoanOriginalAmount,seq(0,1,.125))),
                                           labels=c('0%','12.5%','25%','37.5%','50%','62.5%','75%','87.5%'))

ggplot(aes(y=BorrowerRate,x=LoanOriginalAmountQuantile),data=loanData)+
  geom_boxplot()+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))

ggplot(aes(y=BorrowerRate,x=LoanOriginalAmountQuantile,group=1),
       data=subset(loanData,
                   !is.na(LoanOriginalAmountQuantile)))+
  geom_line(stat='summary',fun.y=median)+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

Smaller loan amounts have a significantly higher median borrower rate.

### BorrowerRate v. DebtToIncomeRatio
```{r echo=FALSE, Bivariate_Plots5,warning=FALSE}
ggplot(aes(y=BorrowerRate,x=log10(DebtToIncomeRatio+1)),
       data=subset(loanData,
                   !is.na(DebtToIncomeRatio)))+
  xlim(0,0.5)+
  geom_point(alpha=1/20)
```

Again, I'm going to cut the DebtToIncomeRatio variable

into quantiles for a cleaner plot...

### Median BorrowerRate v. DebtToIncomeRatio Quantile

```{r echo=FALSE, Bivariate_Plots6,warning=FALSE}
# ggplot(aes(y=BorrowerRate,x=log10(DebtToIncomeRatio)),data=loanData)+
#   geom_line(stat='summary',fun.y=median)

## cut loanData by quantiles...
loanData$DebtToIncomeRatioQuantile <- cut(loanData$DebtToIncomeRatio,
                                          with(subset(loanData,!is.na(DebtToIncomeRatio)),
                                               quantile(DebtToIncomeRatio,seq(0,1,.1))),
                                          labels=c('0%','10%','20%','30%','40%','50%','60%','70%','80%','90%'))

ggplot(aes(y=BorrowerRate,x=DebtToIncomeRatioQuantile),data=loanData)+
  geom_boxplot()

ggplot(aes(y=BorrowerRate,x=DebtToIncomeRatioQuantile,group=1),
       data=subset(loanData,!is.na(DebtToIncomeRatioQuantile)))+
  geom_line(stat='summary',fun.y=median)+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))+
  ylab("Median BorrowerRate")

```


We can see from this graph that borrowers with a higher debt level relative to the others tend to have a higher borrower rate.

### BorrowerRate v. CreditScore

```{r echo=FALSE, Bivariate_Plots8,warning=FALSE}
ggplot(aes(y=BorrowerRate,x=CreditScoreRangeLower),
       data=subset(loanData,!is.na(CreditScoreRangeLower)))+
   geom_jitter(alpha=1/20)+
  xlim(c(440,880))
```

```{r,echo=FALSE,warning=FALSE}
# ggplot(aes(x=LoanOriginationQuarter,y=CreditScoreRangeLower),
#        data=subset(loanData,!is.na(LoanOriginationQuarter)))+
#   geom_boxplot()+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

### BorrowerRate v. CreditGrade
```{r,echo=FALSE,warning=FALSE}
ggplot(aes(y=BorrowerRate,x=CreditGrade),
       data=subset(loanData,!is.na(CreditGrade)))+
  geom_boxplot()
```

Both CreditGrade and CreditScore have exceptionally high correlation to the Borrower Rate

the BorrowerRate. Credit Grade is likely derived from the Credit Score.

This is verified in the plot below.

### Credit Score v. Credit Grade
```{r,echo=FALSE,warning=FALSE}
ggplot(aes(y=CreditScoreRangeLower,x=CreditGrade),
       data=subset(loanData,CreditGrade!=""))+
  geom_boxplot()
```

### Borrower Rate v. Prosper Rating
```{r echo=FALSE, Bivariate_Plots9,warning=FALSE}
ggplot(aes(y=BorrowerRate,x=ProsperRating..Alpha.),
       data=loanData)+
  geom_boxplot()
```

Prosper ratings are even more closely related to the BorrowerRate than credit grades. The Prosper Rating is likely a metric that the Prosper Loan Company uses to asses risk, based off of other parameters.

### BorrowerRate v. StatedMonthlyIncome
```{r echo=FALSE,warning=FALSE}
ggplot(aes(y=BorrowerRate,x=StatedMonthlyIncome),
       data=subset(loanData,
                   IncomeVerifiable=="True"&!
                   is.na(StatedMonthlyIncome)))+
  scale_x_log10(limits=c(500,50000))+
  geom_jitter(alpha=1/20)
```

Once again, I'm going to cut the StatedMonthlyIncome variable

into quantiles for a cleaner plot...

### Median BorrowerRate v. StatedMonthlyIncome Quantile
```{r echo=FALSE,warning=FALSE}
loanData$StatedMonthlyIncomeQuantile<-
  cut(loanData$StatedMonthlyIncome,
      with(subset(loanData,
                  !is.na(StatedMonthlyIncome)),
           quantile(StatedMonthlyIncome,seq(0,1,.1))),
      labels=c('0%','10%','20%','30%','40%','50%','60%','70%','80%','90%'))

ggplot(aes(y=BorrowerRate,x=StatedMonthlyIncomeQuantile,group=1),
       data=subset(loanData,
                   IncomeVerifiable=="True"&!
                   is.na(StatedMonthlyIncomeQuantile)))+
  geom_line(stat='summary',fun.y=median)+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))+
  ylab("Median BorrowerRate")
```

The median BorrowerRate varies consistently with the Quantile of StatedMonthlyIncome.

### BorrowerRate v. IncomeRange
```{r echo=FALSE, Bivariate_Plots11.1,warning=FALSE}
ggplot(aes(y=BorrowerRate,x=IncomeRange),
       data=subset(loanData,
                   IncomeVerifiable=="True"))+
  geom_boxplot()+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

Not surprisingly, there is a similar variation in BorrowerRate amongst Borrowers with different Income Ranges.

```{r echo=FALSE, Bivariate_Plots11,warning=FALSE}
# ggplot(aes(x=IncomeRange,y=StatedMonthlyIncome*12),
#        data=subset(loanData,
#                    IncomeVerifiable=="True"))+
#   geom_boxplot()+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))

# ggplot(aes(x=IncomeRange,y=StatedMonthlyIncome*12),
#        data=subset(loanData,
#                    IncomeVerifiable=="True"))+
#   geom_boxplot()+
#   coord_cartesian(ylim=c(0,200000))+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

### BorrowerRate v. Employment Status
```{r echo=FALSE, Bivariate_Plots14,warning=FALSE}
ggplot(aes(y=BorrowerRate,x=reorder(EmploymentStatus,BorrowerRate,FUN=median)),data=loanData)+
  geom_boxplot()+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))+
  ylab("EmploymentStatus")
```

Unemployed borrowers have a significantly higher median BorrowerRate than the others

### BorrowerRate v. IsBorrowerHomeowner
```{r echo=FALSE, Bivariate_Plots15,warning=FALSE}
ggplot(aes(y=BorrowerRate,x=IsBorrowerHomeowner),
       data=loanData)+
  geom_boxplot()
```

Homeowners have a lower median BorrowerRate than non-homeowners.

### BorrowerRate v. BorrowerState
```{r echo=FALSE, Bivariate_Plots15.873265987230,warning=FALSE,fig.height=8}
ggplot(aes(y=BorrowerRate,x=reorder(BorrowerState, BorrowerRate, FUN=median) ),data=loanData)+
  geom_boxplot()+
  coord_flip()+
  xlab("State")
```

The worst States to get a loan are Maine and Indiana. The Best States are Alabama and North Dakota

### BorrowerRate v. Occupation
```{r echo=FALSE,message=FALSE,warning=FALSE,fig.height=10}
ggplot(aes(y=BorrowerRate,x=reorder(Occupation, BorrowerRate,FUN=median )),data=loanData)+
  geom_boxplot()+
  coord_flip()+
  xlab("Occupation")
```

In General, Higher Paying Occupations with a higher level of education (i.e. Judge, Computer Programmer Engineer) have a lower median borrower rate than lower paying occupations with lower levels of education (i.e. Teacher's Aide, Nurse's Aide, College Freshman). 

# Bivariate Analysis

### -------------------

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?

The average Borrower Rate was decreasing significantly from 2012 to 2014.

The Quantile of DebtToIncomeRatio and Median BorrowerRate are correlated.

CreditScore has a significant correlation to BorrowerRate

CreditGrade is derived from Credit Score

Prosper Ratings are highly related to BorrowerRate.

The Median BorrowerRate goes down as IncomeRange goes up.

StatedMonthlyIncomeQuantile and the Median BorrowerRate are highly correlated.

Unemployed borrowers have a significantly higher Median BorrowerRate than the others.

Homeowners have a lower median BorrowerRate than non-homeowners.

The worst States to get a loan are Maine and Indiana. The Best States are Alabama and North Dakota.

In General, Higher Paying Occupations with a higher level of education (i.e. Judge, Computer Programmer Engineer) have a lower median borrower rate than lower paying occupations with lower levels of education (i.e. Teacher's Aide, Nurse's Aide, College Freshman). 

### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?

CreditGrade is derived from CreditScore.

ProsperRating determines the BorrowerRate almost perfectly, and is likely a score assigned by the prosper loan company based off of other parameters.

### What was the strongest relationship you found?

The strongest Relationship was between BorrowerRate and ProsperRating, however ProsperRating its self is likely composed of other parameters, as noted above. 

Aside from this, the relationship between BorrowerRate and CreditScore was the second strongest.

# Multivariate Plots Section

### ---------------------------


```{r echo=FALSE, Multivariate_Plots11233,message=FALSE,warning=FALSE}
# p1<-ggplot(aes(x=LoanOriginationQuarter,group=1),
#            data=subset(loanData,
#                        !is.na(LoanOriginationQuarter)))+
#   geom_histogram(binwidth=1)+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))
# p2<-ggplot(aes(x=LoanOriginationQuarter,group=1,y=1/BorrowerRate),
#            data=subset(loanData,
#                        !is.na(LoanOriginationQuarter)))+
#   geom_line(stat='summary',fun.y=mean)+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))+
#   ylab("Inverse BorrowerRate")
# p3<-ggplot(aes(x=LoanOriginationQuarter,group=1,y=LoanOriginalAmount),
#            data=subset(loanData,
#                        !is.na(LoanOriginationQuarter)))+
#   geom_line(stat='summary',fun.y=mean)+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))
# grid.arrange(p1,p2,p3,ncol=1)
```

In this section, my main objective is to see how the observed relationships from the previous section hold up and/or change over time.

To start off with, I will investigate the relationship between CreditGrade, Prosper Rating and Credit Score in more detail.

### Median Borrower Rate v. ProsperRating Over Time
```{r echo=FALSE, Multivariate_Plots2.2sdf22,message=FALSE,warning=FALSE}
ggplot(aes(x=LoanOriginationQuarter,group=1,y=BorrowerRate),
       data=subset(loanData,
                   !is.na(LoanOriginationQuarter)&
                     !(ProsperRating..Alpha.=='')))+
  geom_line(aes(group=ProsperRating..Alpha., color=ProsperRating..Alpha.),
            stat="summary",fun.y=mean)+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

### Median Borrower Rate v. CreditGrade Over Time
```{r,echo=FALSE,warning=FALSE}
ggplot(aes(x=LoanOriginationQuarter,group=1,y=BorrowerRate),
       data=subset(loanData,
                   !is.na(LoanOriginationQuarter)&
                     !(CreditGrade=='')))+
  geom_line(aes(group=CreditGrade,color=CreditGrade),
            stat="summary",fun.y=mean)+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

We can see from these two charts that the relationship between the
prosper rating and borrower rate and the relationship between the
credit grade and borrower rate are farely consistent over time.

An unintended, but useful insight from the above visualizations
is that the credit grade data goes up to Q2 2009, 
and the prosper rating data ranges from Q3 2009 onwards.
Given that both parameters are closely related to the BorrowerRate
(as ovserved in the bivariate plots section),
this suggests that the Prosper Loan Company used the borrower's CreditGrade
or credit score up until Q2 2009 as a primary metric in order to determine
the BorrowerRate and used the Prosper Rating thereafter. 

```{r echo=FALSE, Multivariate_Plots3asdfgwe,message=FALSE,warning=FALSE}
loanData$CreditGradeOrProsperRating <-loanData$LoanOriginationQuarter %in% levels(loanData$LoanOriginationQuarter)[0:14]
```

### Borrower Rate v. Credit Score by quarter
```{r echo=FALSE, Multivariate_Plots5sdrge43523,message=FALSE,warning=FALSE}

# p1<-ggplot(aes(y=BorrowerRate,x=CreditScoreRangeLower),
#        data=subset(loanData,
#                    !is.na(LoanOriginationQuarter)&
#                      CreditGradeOrProsperRating))+
#   xlim(c(440,880))+
#   geom_jitter(alpha=1/20)+
#   ggtitle("BorrowerRate v. Credit Score up to Q2 2009")
# 
# p2<-ggplot(aes(y=BorrowerRate,x=CreditScoreRangeLower),
#        data=subset(loanData,
#                    !is.na(LoanOriginationQuarter)&
#                      !CreditGradeOrProsperRating))+
#   xlim(c(440,880))+
#   geom_jitter(alpha=1/20)+
#   ggtitle("BorrowerRate v. Credit Score after Q2 2009")
# grid.arrange(p1,p2)

print(ggplot(aes(y=BorrowerRate,x=CreditScoreRangeLower),
       data=subset(loanData,
                   !is.na(LoanOriginationQuarter)))+
  xlim(c(440,880))+
  geom_jitter(alpha=1/5)+
  facet_wrap(~LoanOriginationQuarter)+
  scale_y_continuous(breaks=c(0.1,0.2,0.3))+
  theme(axis.text.x=element_text(angle = 45, hjust = 1)))
```

### Correlation Between BorrowerRate and CreditScore by Quarter
```{r,echo=FALSE,warning=FALSE}
quarters<- c()
cors1<-c()
for (quarter in levels(loanData$LoanOriginationQuarter)){
  if (quarter!="Q1 2006"&quarter!="Q1 2009"){
    quarters<-c(quarters,quarter)
    cor<-with(subset(loanData,
                       LoanOriginationQuarter==quarter&
                       CreditScoreRangeLower>0),
    cor.test(CreditScoreRangeLower,BorrowerRate))$estimate
    cors1<-c(cors1,cor)
  }
}

creditCor<-data.frame(quarters,cors1)
creditCor$quarters<-factor(creditCor$quarters,levels=quarters)
ggplot(aes(x=quarters,group=1,y=cors1),data=creditCor)+geom_line()+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))+
  xlab("LoanOriginationQuarter")+
  ylab("Correlation Coefficient")
```

The correlation between credit score and borrower rate varies significantly over time. This means it may be useful to model each quarter separately, or make models for different segments of time.

In particular, Correlation is higher Goes down after Q3 2011.

### LoanOriginalAmount by Quarter
```{r echo=FALSE, Multivariate_Plots62353245,message=FALSE,warning=FALSE}
# ggplot(aes(x=LoanOriginalAmount,y=BorrowerRate),data=loanData)+
#   geom_point()+
#   facet_wrap(~LoanOriginationQuarter,ncol=4)+
#   scale_y_continuous(limits=c(0.2,0.4))

# ggplot(aes(x=LoanOriginalAmountQuantile,group=1,y=BorrowerRate),
#        data=subset(loanData,!is.na(LoanOriginalAmountQuantile)))+
#   facet_wrap(~LoanOriginationQuarter,ncol=5)+
#   geom_line(stat='summary',fun.y=median,color='red')+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))

# p1<-ggplot(aes(x=LoanOriginalAmount,y=BorrowerRate),
#            data=subset(loanData,CreditGradeOrProsperRating==TRUE))+
#   geom_point(aes(color=CreditGrade))+
#   facet_wrap(~Term,ncol=1)+
#   scale_color_brewer(~CreditGrade)+
#   ggtitle("Pre-2009")
# 
# p2<-ggplot(aes(x=LoanOriginalAmount,y=BorrowerRate),
#            data=subset(loanData,CreditGradeOrProsperRating==FALSE))+
#   geom_point(aes(color=ProsperRating..Alpha.))+
#   facet_wrap(~Term,ncol=1)+
#   scale_color_brewer(~ProsperRating..Alpha.)+
#   ggtitle("Post-2009")
# 
# grid.arrange(p1,p2,ncol=1)

ggplot(aes(x=LoanOriginalAmount,y=BorrowerRate),
           data=subset(loanData,
                         !is.na(LoanOriginationQuarter)&
                         Term=="36"))+
  geom_point(alpha=1/3)+
  facet_wrap(~LoanOriginationQuarter)
```

LoanOriginalAmount appears to play a role in the BorrowerRate at least up until Q2 2010. 

```{r echo=FALSE, Multivariate_Plots734236,message=FALSE,warning=FALSE}
# ggplot(aes(x=LoanOriginalAmount,y=BorrowerRate),data=subset(loanData,CreditGradeOrProsperRating))+
#   geom_point(aes(color=CreditGrade))+
#   facet_wrap(~LoanOriginationQuarter)+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))+
#   scale_color_brewer(~CreditGrade)
```

```{r,echo=FALSE,warning=FALSE}
# ggplot(aes(x=LoanOriginalAmount,group=1,y=BorrowerRate),
#        data=subset(loanData,CreditGrade!=""))+
#   geom_point()+
#   facet_wrap(~CreditGrade)+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))

# ggplot(aes(x=LoanOriginalAmountQuantile,group=1,y=BorrowerRate),
#        data=subset(loanData,CreditGradeOrProsperRating==TRUE))+
#   facet_wrap(~LoanOriginationQuarter)+
#   geom_line(aes(group=CreditGrade,color=CreditGrade),stat='summary',fun.y=median)+
#   ylab("Mean BorrowerRate")+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))


## do a cor.test by credit grade here, when I figure out how to do that...
```

### Correlation of BorrowerRate and LoanOriginalAmount by Quarter
```{r,echo=FALSE,warning=FALSE}
quarters<- c()
cors2<-c()
for (quarter in levels(loanData$LoanOriginationQuarter)){
  if (quarter!="Q1 2006"&quarter!="Q1 2009"){
    quarters<-c(quarters,quarter)
    cor<-with(subset(loanData,
                       LoanOriginationQuarter==quarter),
    cor.test(CreditScoreRangeLower,LoanOriginalAmount))$estimate
    cors2<-c(cors2,cor)
  }
}

LoanAmountCor<-data.frame(quarters,cors2)
LoanAmountCor$quarters<-factor(creditCor$quarters,levels=quarters)
ggplot(aes(x=quarters,group=1,y=cors2),data=LoanAmountCor)+geom_line()+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))+
  xlab("LoanOriginationQuarter")+
  ylab("Correlation Coefficient")
```

Correlation between BorrowerRate and LoanOriginalAmount goes up in 2007 and back down around Q2 2013.


```{r echo=FALSE, message=FALSE,warning=FALSE}
# ggplot(aes(x=LoanOriginalAmount,y=BorrowerRate),
#        data=subset(loanData,CreditGradeOrProsperRating==FALSE&Term=='36'))+
#   geom_point(aes(color=ProsperRating..Alpha.))+
#   facet_wrap(~LoanOriginationQuarter)+
#   scale_color_brewer(~ProsperRating..Alpha.)
```

```{r echo=FALSE, Multivariate_Plots8.1276348762,message=FALSE,warning=FALSE}
# ggplot(aes(x=LoanOriginalAmount,y=BorrowerRate),
#        data=subset(loanData,LoanOriginationQuarter %in% levels(LoanOriginationQuarter)[15:20]&Term=='36'&!(ProsperRating..Alpha.=="")))+
#   geom_point(aes(color=ProsperRating..Alpha.))+
#   facet_wrap(~LoanOriginationQuarter)+
#   scale_color_brewer(~ProsperRating..Alpha.)
```

```{r,echo=FALSE}
# ggplot(aes(x=LoanOriginalAmount,y=BorrowerRate),
#        data=subset(loanData,
#                    LoanOriginationQuarter %in%
#                      levels(LoanOriginationQuarter)[15:19]&
#                      Term=='36'&
#                      ProsperRating..Alpha.!=''))+
#   geom_point(alpha=1/5)+
#   facet_wrap(~ProsperRating..Alpha.)
```

```{r echo=FALSE, Multivariate_Plots8.12fdrew55sgtw,message=FALSE,warning=FALSE}
# ggplot(aes(x=LoanOriginalAmount,y=BorrowerRate),
#        data=subset(loanData,CreditGradeOrProsperRating==FALSE&Term=='12'))+
#   geom_point(aes(color=ProsperRating..Alpha.))+
#   facet_wrap(~LoanOriginationQuarter)
```

```{r echo=FALSE, Multivariate_Plots8.132465322fdsgtw,message=FALSE,warning=FALSE}
# ggplot(aes(x=LoanOriginalAmount,y=BorrowerRate),
#        data=subset(loanData,CreditGradeOrProsperRating==FALSE&Term=='60'))+
#   geom_point(aes(color=ProsperRating..Alpha.))+
#   facet_wrap(~LoanOriginationQuarter)
```

### Median Borrower Rate v. Income Level by quarter.
```{r echo=FALSE, multivariate_plots.726345341876321895876,message=FALSE,warning=FALSE}
ggplot(aes(x=LoanOriginationQuarter,group=1,y=BorrowerRate),
       data=subset(loanData,
                   !is.na(LoanOriginationQuarter)&
                     IncomeRange%in%levels(IncomeRange)[1:5]))+
  geom_line(aes(group=IncomeRange,color=IncomeRange),stat="summary",fun.y=median)+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

The relationship between IncomeRange and BorrowerRate appears to be much stronger after 2009. In order to get a clearer picture of this, The StatedMonthlyIncomeQuantile is plotted against the median borrower rate by quarter below.   

### Median BorrowerRate v. StatedMonthlyIncome Quantile
```{r echo=FALSE, multivariate_plots.ar324532ewur,message=FALSE,warning=FALSE}
# ggplot(aes(x=log10(StatedMonthlyIncome),y=BorrowerRate),
#        data=subset(loanData,!is.na(LoanOriginationQuarter)&IncomeRange%in%levels(IncomeRange)[1:4]))+
#   geom_point()+
#   facet_wrap(~LoanOriginationQuarter)+
#    theme(axis.text.x=element_text(angle = 45, hjust = 1))

ggplot(aes(x=StatedMonthlyIncomeQuantile,group=1,y=BorrowerRate),
       data=subset(loanData,!is.na(LoanOriginationQuarter)&IncomeRange%in%levels(IncomeRange)[1:4]))+
  geom_line(stat="summary",fun.y=median)+
  facet_wrap(~LoanOriginationQuarter)+
  ylab("Median BorrowerRate")+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))
```

It looks like Income is more closely correlated to the BorrowerRate after Q2 2009. 
This is verified in the chart below.

### Correlation Between BorrowerRate and log10(StatedMonthlyIncome+1) by Quarter.
```{r,echo=FALSE,warning=FALSE}
quarters<- c()
cors4<-c()
for (quarter in levels(loanData$LoanOriginationQuarter)){
  if (quarter!="Q1 2006"&quarter!="Q1 2009"){
    quarters<-c(quarters,quarter)
    cor<-with(subset(loanData,
                       LoanOriginationQuarter==quarter),
    cor.test(CreditScoreRangeLower,StatedMonthlyIncome+1))$estimate
    cors4<-c(cors4,cor)
  }
}

IncomeCor<-data.frame(quarters,cors4)
IncomeCor$quarters<-factor(creditCor$quarters,levels=quarters)
ggplot(aes(x=quarters,group=1,y=cors4),data=IncomeCor)+geom_line()+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))+
  xlab("LoanOriginationQuarter")+
  ylab("Correlation Coefficient")
```
Correlation between StatedMonthlyIncome and BorrowerRate falls in Q2 2012


```{r echo=FALSE, multivariate_plots.asf325iy32y7y,message=FALSE,warning=FALSE}
# ggplot(aes(x=CreditGrade,y=StatedMonthlyIncome),
# data=subset(loanData,CreditGradeOrProsperRating==TRUE))+
#   geom_boxplot()+
#   coord_cartesian(ylim=c(500,10000))
# 
# ggplot(aes(x=ProsperRating..Alpha.,y=StatedMonthlyIncome),
# data=subset(loanData,CreditGradeOrProsperRating==FALSE))+
#   geom_boxplot()+
#   coord_cartesian(ylim=c(500,10000))
```

Next, we know from the previous DebtToIncomRatioQuantile v. Median BorrowerRate line plot that BorrowerRate and DebToIncomeRatio are related. Lets try and find where those relationships are particularly strong.

### BorrowerRate v. log transform of DebtToIncomeRatio by Quarter with constant credit score of 760
```{r,echo=FALSE,warning=FALSE}
ggplot(aes(y=BorrowerRate,x=log10(DebtToIncomeRatio+1)),data=
         subset(loanData,
                !is.na(DebtToIncomeRatio)&
                  !(is.na(LoanOriginationQuarter))&
                  CreditScoreRangeLower==760))+
  geom_point(alpha=1/5)+
  facet_wrap(~LoanOriginationQuarter)+
  coord_cartesian(xlim=c(0.001,.25))
```

### Correlation of BorrowerRate and log10(DebtToIncomeRatio+1) by Quarter
```{r,echo=FALSE,warning=FALSE}
# for (i in levels(factor(loanData$LoanOriginationQuarter))) {
#   print(i)
#   if (i!="Q2 2009"){
#     print(with(subset(loanData,
#          (LoanOriginationQuarter==i)&
#          DebtToIncomeRatio<.5),
#     cor.test(BorrowerRate,DebtToIncomeRatio))["estimate"])
#   }
# }

quarters<- c()
cors3<-c()
for (quarter in levels(loanData$LoanOriginationQuarter)){
  if (quarter!="Q1 2006"&quarter!="Q1 2009"){
    quarters<-c(quarters,quarter)
    cor<-with(subset(loanData,
                       LoanOriginationQuarter==quarter),
    cor.test(CreditScoreRangeLower,log10(DebtToIncomeRatio+1)))$estimate
    cors3<-c(cors3,cor)
  }
}

DebtCor<-data.frame(quarters,cors3)
DebtCor$quarters<-factor(creditCor$quarters,levels=quarters)
ggplot(aes(x=quarters,group=1,y=cors3),data=DebtCor)+geom_line()+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))+
  xlab("LoanOriginationQuarter")+
  ylab("Correlation Coefficient")
```

Correlation Between BorrowerRate and DebtToIncomeRatio is generally low, but goes up in Q2 2009, and back down in Q2 2011. 

There appears to be a consistent pattern here. There seems to be a change around Q2 2009, when the prosper rating went into affect, and around Q2 2011 in the methodology used to calculate the loans.

```{r,echo=FALSE,warning=FALSE}
# ggplot(aes(y=BorrowerRate,x=log10(DebtToIncomeRatio+1)),data=
#          subset(loanData,
#                 !is.na(DebtToIncomeRatio)&
#                   ProsperRating..Alpha.!=''&
#                   !CreditGradeOrProsperRating))+
#   geom_point(alpha=1/15,aes(color=ProsperRating..Alpha.))+
#   facet_wrap(~LoanOriginationQuarter)
```

```{r,echo=FALSE,warning=FALSE}
# for (i in levels(factor(loanData$CreditScoreRangeLower))) {
#   print(i)
#   if ((i>"600")){
#     print(with(subset(loanData,
#          (CreditScoreRangeLower==i)&
#          substr(LoanOriginationQuarter,6,7)>"09"&
#          DebtToIncomeRatio<.5),
#     cor.test(BorrowerRate,DebtToIncomeRatio+1))["estimate"])
#   }
# }
```

```{r echo=FALSE, Multivariate_Plots18532ssagewgdsffsd78yhuf,message=FALSE,warning=FALSE}
# loanData$PercentDebt <- with(loanData,
#                                         LoanOriginalAmount/(DebtToIncomeRatio)*StatedMonthlyIncome)
# 
# summary(loanData$PercentDebt)
# 
# ggplot(aes(x=PercentDebt),data=
#          subset(loanData,
#                 DebtToIncomeRatio<0.5))+
#   geom_histogram()+
#   scale_x_log10(limits=c(150000,1000000000))
```

```{r,echo=FALSE,warning=FALSE}
# ggplot(aes(y=BorrowerRate,x=PercentDebt),
#        data=subset(loanData,
#                    DebtToIncomeRatio<.5))+
#   geom_jitter(alpha=1/5)+
#   scale_x_log10(limits=c(10000000,1000000000))+
#   facet_wrap(~LoanOriginationQuarter)
```

```{r,echo=FALSE,warning=FALSE}
# for (i in levels(factor(loanData$LoanOriginationQuarter))) {
#   print(i)
#   if (i!="Q2 2009"){
#     print(with(subset(loanData,
#          (LoanOriginationQuarter==i)&
#         CreditScoreRangeLower==760&
#          DebtToIncomeRatio<.5),
#     cor.test(BorrowerRate,log10(PercentDebt)))["estimate"])
#   }
# }
```


# Linear Models
###------------

### Linear Model for all Loans
```{r echo=FALSE,message=FALSE,warning=FALSE}
#linear regression earlier data credit score 600 or better with loan amount and credit score...
# fit1<-with(
#   subset(
#     loanData,
#     (CreditGradeOrProsperRating==TRUE)&
#       CreditScoreRangeLower>600),
#      lm(BorrowerRate~
#           LoanOriginalAmount+
#           (CreditScoreRangeLower)))
# 
# summary(fit1)
# 
# ggplot(aes(y=BorrowerRate,x=CreditScoreRangeLower*-6.846e-04+LoanOriginalAmount*2.284e-06),
#        data=subset(
#     loanData,
#     (CreditGradeOrProsperRating==TRUE)&CreditScoreRangeLower>600&IncomeVerifiable=="True"))+
#   geom_jitter(alpha=1/5)+
#   geom_smooth(method="lm")


fit1<-with(
  subset(
    loanData,
      CreditScoreRangeLower>0&
      IncomeVerifiable=="True"),
     lm(BorrowerRate~
          LoanOriginalAmount+
          CreditScoreRangeLower+
          log10(DebtToIncomeRatio+1)))

summary(fit1)
```

The R^2 value for this fit is quite low. 

Given the variation in the median borrower rate over time, as well as well as the variation over time amongst several variables in the correlation to the borrower rate, there may be better results if models are created for different periods of time.

We know the following from the previous plots:

Correlation between CreditScoreRangeLower and BorrowerRate goes down around Q1-Q2 2011.

Correlation between LoanOriginalAmount and BorrowerRate goes down around Q1-Q2 2011.

Correlation between StatedMonthlyIncome and BorrowerRate goes down around Q1 2011

Correlation between DebtToIncomeRatio and BorrowerRate goes up in Q2 2009 and back down in Q1 2011.

As such, it seems like good intervals to split the data would be from the beginning to Q2 2009, and from Q3 2009 through Q1 2011. 

### Model For loans in or before Q2 2009
```{r echo=FALSE,message=FALSE,warning=FALSE}
fit2<-with(
  subset(
    loanData,
    (CreditGradeOrProsperRating)&
      CreditScoreRangeLower>0&
      IncomeVerifiable=="True"),
     lm(BorrowerRate~
          LoanOriginalAmount+
          CreditScoreRangeLower))

summary(fit2)

ggplot(aes(y=BorrowerRate,
           x=6.083e-01+
             CreditScoreRangeLower*-6.846e-04+
             LoanOriginalAmount*2.284e-06),
       data=subset(
    loanData,
    (CreditGradeOrProsperRating==TRUE)&CreditScoreRangeLower>0&IncomeVerifiable=="True"))+
  geom_jitter(alpha=1/5)+
  geom_smooth(method="lm")
```

### Model for Loans Originating from Q3 2009 through Q1 2011
```{r echo=FALSE,message=FALSE,warning=FALSE}
fit3<-with(
  subset(
    loanData,
    LoanOriginationQuarter %in% levels(loanData$LoanOriginationQuarter)[15:21]&
      Term=="36"),
     lm(BorrowerRate~
          LoanOriginalAmount+
          log10(DebtToIncomeRatio+1)+
          CreditScoreRangeLower))

summary(fit3)

ggplot(aes(y=BorrowerRate,
           x=1.020e+00+ 
             LoanOriginalAmount*1.661e-06+
             log10(DebtToIncomeRatio + 1)*1.471e-01+
             CreditScoreRangeLower*-1.182e-03),
       data=subset(
    loanData,
      LoanOriginationQuarter %in% levels(loanData$LoanOriginationQuarter)[15:21]&
      Term=="36"))+
  geom_jitter(alpha=1)+
  geom_smooth(method="lm")+
  coord_cartesian(xlim=c(0,0.35))
```

Much better models resulted from the division of loans into separate periods. The log transform of DebtToIncomeRatio was included in the second model, but not in the first, since it did not improve the model. This makes sense, given that we saw that the correlation between DebtToIncomeRatio and BorrowerRate is generally higher after Q2 2009.

Both models benefited from LoanOriginalAmount as expected.

The inclusion of Stated Monthly Income did not have an effect on either model, so it was left out. This means that the observed relationship may have had more to do with generally better CreditScore or DebtToIncomeRatio for borrowers with higher income.


# Multivariate Analysis

### ---------------------

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?

The average loan amount is closely correlated with the number of loans given out.

Credit score is correlated to Borrower Rate, although that correlation changes by quarter.

As of 2009, there is a strict cutoff at a credit score of 600.

There isn't much variation in Borrower Rate by credit Score for those borrowers with a credit score of less than 600.

The correlation between the borrower rate and the loan amount becomes higher as the CreditGrade goes up.

The correlation between the borrower rate and the loan amount becomes higher as the Prosper Rating goes up.

There is a correlation between loan amount and borrower rate exists up until 2011.

The median borrower rate is related to IncomeRange.

The trend towards lower BorrowerRate for higher StatedMonthlyIncome is consistent accross all of the quarters.

The trend towards higher BorrowerRate for Higher DebtToIncomeRatio is consistent accross all of the quarters.

The quantile Of Available Monthly income is correlated to the median Borrower Rate.

### Were there any interesting or surprising interactions between features?

CreditGrades were only used up until 2009, after which ProsperRatings were primarily used to determine BorrowerRate.

StatedMonthlyIncome Varies significantly more accross different ProsperRatings than it does accross different CrediGrades.

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

#### I creted the following linear models, in order.

(The models apply to a subset of the data taken from 2006 to 2009 with a CreditScoreRangeLower of 600 or less.)

##### Using CreditScoreRangeLower and LoanOriginalAmount for all loans
R^2:0.2746

##### Using CreditScoreRangeLower and LoanOriginalAmount for loans originating before 2009
R^2: 0.5189

##### Using CreditScoreRangeLower, LoanOriginalAmount and log10(DebtToIncomeRatio+1) for loans originating from Q3 2009 through Q2 2011
R^2: 0.4953

The first model is the most general one, but has a very low R^2 value, and therefore does not represent the data very well.

The Second two models, with higher R^2 values, do a much better job of modelling the data. Niether model however is capable of taking into account the fluctuations in the BorrowerRate over time due to economic trends.


# Final Plots and Summary
### ----------------------------
### Plot One, BorrowerRate v. Occupation
```{r echo=FALSE, Plot_One,message=FALSE,warning=FALSE,fig.height=10}
### OLD PLOT...
###
###

# p1<-ggplot(aes(x=LoanOriginationQuarter,group=1),data=loanData)+
#   geom_histogram(binwidth=1)+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))
# p2<-ggplot(aes(x=LoanOriginationQuarter,group=1,y=1/BorrowerRate),data=loanData)+
#   geom_line(stat='summary',fun.y=mean)+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))
# p3<-ggplot(aes(x=LoanOriginationQuarter,group=1,y=LoanOriginalAmount),data=loanData)+
#   geom_line(stat='summary',fun.y=mean)+
#   coord_cartesian(ylim=c(0,20000))+
#   theme(axis.text.x=element_text(angle = 45, hjust = 1))
# 
# grid.arrange(p1,p2,p3,ncol=1)


# 
# x <- levels(loanData$LoanOriginationQuarter)
# 
# d3 <- data.frame(x=loanData$LoanOriginationQuarter, y=loanData$LoanOriginalAmount)
# xy <- expand.grid(x=x, y=x)
# d2 <- data.frame(x=loanData$LoanOriginationQuarter, y=loanData$BorrowerRate)
# d1 <- data.frame(x=loanData$LoanOriginationQuarter, y=loanData$BorrowerRate)
# 
# d1$panel <- "Mean Loan Amount"
# d2$panel <- "Mean Inverse BorrowerRate"
# d3$panel <- "Number of Loans"
# 
# d <- rbind(d1, d2,d3)
# 
# p <- ggplot(data = loanData, mapping = aes())
# p <- p + facet_grid(panel~., scale="free")
# p <- p + layer(data=d3, mapping=aes(x = loanData$LoanOriginationQuarter,group=1, y = ..count..),  geom =c("bar"))
# p <- p + layer(data= d1,  mapping=aes(x = loanData$LoanOriginationQuarter,group=1, y = loanData$LoanOriginalAmount),geom = c( "line"), stat = "summary",fun.y = mean)
# p <- p + layer(data=d2, mapping=aes(x = loanData$LoanOriginationQuarter,group=1, y = 1/loanData$BorrowerRate),  geom =c("line"), stat = "summary",fun.y=mean)
# p<-p+  theme(axis.text.x=element_text(angle = 45, hjust = 1))
# p

###NEW PLOT
ggplot(aes(y=BorrowerRate,x=reorder(Occupation, BorrowerRate,FUN=median )),data=loanData)+
  geom_boxplot()+
  coord_flip()
```

### Description One
This plot shows borrower occupation on the y axis in order of median borrower rate, with highest median borrower rate at the top. BorrowerRate is shown on the x axis.

In General, Higher Paying Occupations with a higher level of education (i.e. Judge, Computer Programmer, Engineer,Professor) have a lower median borrower rate than do lower paying occupations with lower levels of education (i.e. Teacher's Aide, Nurse's Aide, College Freshman,Clerical,Bus Driver).

While it is not clear weather the borrower's occupation contributes directly to the borrower rate, it is interesting to see how the general prominance of the Borrower's occupation relates to the BorrowerRate.

### Plot Two
```{r echo=FALSE,warning=FALSE}
ggplot(aes(x=LoanOriginationQuarter,y=BorrowerRate,group=1),
       data=subset(loanData,
                   !is.na(LoanOriginationQuarter)))+
  geom_line(stat='summary',fun.y=median)+
  theme(axis.text.x=element_text(angle = 45, hjust = 1))+
  ylab("Median BorrowerRate")
```

### Description Two

This plot shows the median BorrowerRate, by quarter.

We can see from this plot that the mean borrower rate rises in 2010 and falls in 2013. More Generally, this plot demonstrates the variable nature of the BorrowerRate.

Given the relatively large number of loans, one might otherwise expect the median borrower rate to remain relatively stagnent relative to the rest of the loans over time. This variation means that the borrower rate is subject to economic trends over time in addition to statistics of the borrower and the original loan terms.

### Plot Three
```{r echo=FALSE,Plot_Three,message=FALSE,warning=FALSE}
p1<- ggplot(aes(y=BorrowerRate,
           x=6.083e-01+
             CreditScoreRangeLower*-6.846e-04+
             LoanOriginalAmount*2.284e-06),
       data=subset(
    loanData,
    (CreditGradeOrProsperRating==TRUE)&
  CreditScoreRangeLower>600))+
  geom_jitter(alpha=1/5)+
  geom_smooth(method="lm")+
  ggtitle("Linear Model for Loans Originating from Q1 2006 through Q2 2009")


p2<-ggplot(aes(y=BorrowerRate,
           x=1.020e+00+ 
             LoanOriginalAmount*1.661e-06+
             log10(DebtToIncomeRatio + 1)*1.471e-01+
             CreditScoreRangeLower*-1.182e-03),
       data=subset(
    loanData,
      LoanOriginationQuarter %in% levels(loanData$LoanOriginationQuarter)[15:21]&
      Term=="36"))+
  geom_jitter(alpha=1/2)+
  geom_smooth(method="lm")+
  coord_cartesian(xlim=c(0,0.35))+
  ggtitle("Linear Model for Loans Originating from Q3 2009 through Q1 2011")

grid.arrange(p1,p2)
```

### Description Three

The above plots show two linear model of the Borrower Rate, the first for loans up to Q2 2009, and the second for loans from Q3 2009 through Q1 2011.

This shows that variation in the BorrowerRate from Q1 2006 to Q2 2009 is explained primarily by the LoanOriginalAmount and the Credit Score for the borrowers with a credit score of greater than 0. Additionally, variation in the BorrowerRate from Q3 2009 through Q1 2011 is explained primarily by the LoanOriginalAmount, the Credit Score and the DebtToIncomeRatio.

###### The fact that DebtToIncomeRatio played a role in linear model for the second period, but not in the first is interesting. According to the National Bureau of Economic Research, the recession ended in the second quarter of 2009, which is also the end of the first model and start of the second. This may imply that the ability of the borrower to pay the loan was taken more seriously after the recession.
------

# Reflection


The prosper loan data set contains information on more than 100,000 loans from 2006 to 2014. My objective was to explore trends and factors contributing to the borrower rate. I started out by gaining an understanding of each of the variables in the investigation, and then went on to investigate the relationship between each variable and the borrower rate. I eventually explored the strongest relationships in further detail in order to get a better sense of when and where these relationships were strongest.Finally, I created a linear model using LoanOriginalAmount and CreditScoreRangeLower. This linear was insufficient, so I created separate linear models for loans before Q2 2009 and from Q3 2009 through Q1 2011.

The biggest struggle was working out how the different variables related to the borrowerRate over time. For example, DebtToIncomeRatio did not appear to be correlated at all to the BorrowerRate, until the correlation was taken by quarter. Another example is that StatedMonthlyIncome Seemed to have a relationship to BorrowerRate, but was found to be insignificant after it was finally applied in the linear model.

The model is limited primarily by the small number of parameters used in the investigation. The data set contains some 42 parameters, of which several (in addition to those used) may have some influence on the BorrowerRate including BorrowerRateCategory, Reccomendations, Investment from friends, and Investors. The model is also limited by various fluctuations in time in the median borrowerRate over time. Therefore, further investigation could use a wider array of variables in order to try to get a better model of the borrower rate. Further investigation may also take into account contextual economic data in order to account for the general fluctuations in BorrowerRate over time. For example, I could separate the data by borrowerRateCategory, and try to explain the variation within each category, or I might get data on the national interest rate, and use that as a parameter or multiplyer in the models.